---
title: "Project - STAT 151A"
author: "Aditya Jhanwar"
date: "12/8/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
setwd("~/151A/Project/")
rm(list=ls())
library(tidyverse)
```

```{r load_data}
load(url("http://www.stat.berkeley.edu/users/nolan/data/baseball2012.rda"))
baseball = as_tibble(baseball)
```

# Data Exploration and Feature Creation 

### 1)

The first step is to clean the `baseball` data by removing unecessary explanatory variables and entries missing a `salary (observed Yi)` value.

```{r cleaning.data}
baseball = baseball %>% select(-c("ID", "yearID", "teamID", "lgID", "nameFirst", "nameLast", "G_batting")) # remove unecessary variables
baseball = baseball %>% drop_na(salary) # remove units with no salary values
```

Next, I followed the author's process in creating new features as decribed in the textbook.

```{r derived.vars}
baseball = baseball %>% mutate(AVG = CH/CAB,
                               OBP = 100*(CH + CBB)/(CAB + CBB),
                               CAB.avg  = CAB/years,
                               CH.avg   = CH/years,
                               CHR.avg  = CHR/years,
                               CR.avg   = CR/years,
                               CRBI.avg = CRBI/years)
```

Finally, I cleaned the `Position` and `Years` explanatory variables through reimpementing them as dummy variables.

```{r dummy.vars.position}
new.pos = c()
MI = c("12", "1S", "23", "2B", "2S", "3S", "O2", "SS")
C  = c("C", "C1", "OC")
CF = c("CF")
  
# Assign new factor assignment for MI (middle infielders), C (catcher), CF (center field), and O (other)
for (i in baseball$POS){
  if      (i %in% MI) { new.pos = c(new.pos, "MI") }
  else if (i %in% C)  { new.pos = c(new.pos, "C")  }
  else if (i %in% CF) { new.pos = c(new.pos, "CF") }
  else                { new.pos = c(new.pos, "O")  }
}

baseball$POS = relevel(factor(new.pos), "O")
```

```{r dummy.vars.salary}
new.years = c()
neg.cont  = 6
neg.sal   = 3

for (i in baseball$years){
  if      (i >= neg.cont) { new.years = c(new.years, "negotiate.contract") }
  else if (i >= neg.sal)  { new.years = c(new.years, "negotiate.salary")   }
  else                    { new.years = c(new.years, "other")              }
}

baseball$years = relevel(factor(new.years), "other")
```

```{r}
lm.fit = lm(log(salary) ~ ., data = baseball)
new.baseball = as_tibble(model.matrix(lm.fit)[,-1])
new.baseball$salary = baseball$salary
```

\newpage

### 3)

```{r}

```

\newpage

\newpage

# Data Analysis 

### 1)

```{r}

```


























